### 2022-02-28 paper.tex
added smoothng pics

The starting point in scRNA-Seq is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat or ScanPy, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors  of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components. Therefore, Euclidean distances between these "PCA space" vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data. The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

@@

As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described. Figure 1a shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

@@

\begin{figure}
\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
\caption{\textbf{Effect of read position:} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells.}
\end{figure}

@@

We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval. In Figure \ref{}, such an interval is covered by a single read for each of the two cells shown. The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonous interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region. The reason for this is that we reduce the variation in situation as the one depicted in FIgure \ref{}, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.

How should one obtain the ensemble average (the curved line in Fig.\ \ref{})? A simple approach to get a value for a spcific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated. However, especially when only few cells offer coverage, these averages will be very noisy. Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood. The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with XXX bp and used this value for the examples presented here.

@@

@@

Therefore, we propose the folloing approach: Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{}. Then, calculate for each window the variance of these values over all cells. Select, say, the top XX\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR. Then, calculate for each of these laregr VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.

@@

\appendix

\section{Appendix: Methods in Detail}

@@

\bibliography{scbs}

>>>>

The starting point in scRNA-Seq is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. As a first step towards assigning cell types or states to cells, one needs to establish which cells are similar to each other, and to this end, find a way to quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-5}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors  of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components. Therefore, Euclidean distances between these "PCA space" vectors provide a robust dissimilarity score, which can subsequently be used as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data. The PCA space representation is furthermore used for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

\begin{figure}
	\framebox[\columnwidth]{
		\begin{minipage}[c][3.5cm]{\columnwidth}
			\centering Figure
		\end{minipage}
	}
	\caption{\small UMAP representations of the single-cell bisulfite data of XXX et al., calculated \textbf{(A)} using conventional preprocessing and \textbf{(B)} using the approach described here.}
	\label{umaps}
\end{figure}

@@

As an example, we will us in this paper data from XX et al., who have performed sc-BS on [...] and analysed their data in the manner just described. Figure \ref{umaps}A shows a UMAP plot, derived from our re-analysis of the data, using a tiling of the genome into windows of size 100 kb. As one can see, the approach is suitable to see clear structure in the data, observe ...

@@

\begin{figure}
	A\\
	\includegraphics[width=\columnwidth]{need_for_relative_meth.png}
	B\\
	.\hspace{.39cm}\includegraphics[width=.83\columnwidth]{smoothing_residuals.png}
\caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
\label{smoothres}
\end{figure}

@@

We first discuss the task of quantifying the level of methylation in given, fixed, genomic interval. In Figure \ref{smoothres}A, such an interval is covered by a single read for each of the two cells shown. The read from the cell 2 shows much more methylation than the read from cell 1, and a simple analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonous interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

@@

The signal to noise ratio in a matrix thus obtained will be much better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cell's covered CpG in a region. The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{smotthres}A, where absolute methylation might differ strongly even though there is no actual evidence for a difference between the two cells.

How should one obtain the ensemble average (the curved line in Fig.\ \ref{smoothres}B)? A simple approach to get a value for a spcific CpG would be to take all cells with read coverage for the CpG and use the fraction of these that show the CpG as methylated. However, especially when only few cells offer coverage, these averages will be very noisy. Therefore, we propose to smoothen using a kernel smoother, i.e., by performing a kernel-weighted average over the CpG site's neighborhood. The kernel bandwidth (i.e., the size of the neighborhood to average over) is a tuning parameter; we got good results with XXX bp and used this value for the examples presented here.

@@

\begin{figure}
	\includegraphics[width=\columnwidth]{vmr.png}
	\caption{\small \textbf{Finding variably methylated regions (VMR):} The chromosomes are divided up into overlapping windows (first four shown at the bottom), and for each window, the cells' methylation values are calculated as described and as depicted in Fig.\ \ref{smoothres}B. Then, the variance of these values are calculated (each point represents one of the overlapping windows), a threshold (red line) is chosen such that a chosen quantile of windows have a variance exceeding the threshold. Windows with above-threshold variance are merged if they overlap, yielding the ``variably methylated regions''.}
	\label{vmr}
\end{figure}

@@

Therefore, we propose the following approach (Fig. \ref{vmr}): Divide up the chromosome into many \emph{overlapping} windows, that start at regular multiples of a fixed, small, step size. Quantify the methylation of each cell in each window by averaging the cell's methylation residuals over all CpGs in the window, as described above and depicted in Figure\ \ref{smoothres}B. Then, calculate for each window the variance of these values over all cells. Select, say, the top XX\% windows with the highest variances and mark them as VMRs. Wherever thus marked windows overlap or are divided by only a small gap, merge them into one larger VMR. Then, calculate for each of these laregr VMRs the methylation signal, as before, by averaging for each cell over the residuals of all contained CpGs.

@@

\section{Methods}

@@

{\small \bibliography{scbs}}
